---
title: "FinalProject"
author: "Bailey Allord, Laura Sanchez, Valentina Ramirez"
subtitle: MGSC 310 Final Project
output:
  html_document:
    df_print: paged
  html_notebook: default
---

```{r setup, include=FALSE}
library(knitr)

# Change the number in set seed to your own favorite number
set.seed(1818)
options(width=70)
options(scipen=99)


# this sets text outputted in code chunks to small
opts_chunk$set(tidy.opts=list(width.wrap=50),tidy=TRUE, size = "vsmall")  
opts_chunk$set(message = FALSE,                                          
               warning = FALSE,
               # "caching" stores objects in code chunks and only rewrites if you change things
               cache = FALSE,                               
               # automatically downloads dependency files
               autodep = TRUE,
               # 
               cache.comments = FALSE,
               # 
               collapse = TRUE,
               # change fig.width and fig.height to change the code height and width by default
               fig.width = 5.5,  
               fig.height = 4.5,
               fig.align='center')
```

```{r setup-2}
sessionInfo()
getwd()
```

```{r setup-3}
library('tidyverse')
library('here')
library('dplyr')
library('magrittr')
library('partykit')
library('ggplot2')
library('ISLR')
library('NbClust')
library('yardstick')
library('plotROC')
library('rsample')
library('splitstackshape')
library('DescTools')
library('randomForest')
library('randomForestExplainer')
library('sjPlot')
```


## Load data

```{r}
elec <- read.csv(here("Final_BaileyValeLaura", "election_votes.csv"))
pov <- read.csv(here("Final_BaileyValeLaura", "poverty_income.csv"))
unemp <- read.csv(here("Final_BaileyValeLaura", "unemployment.csv"))
```


## Data transformation

```{r}
dem_elec <- elec %>% 
  filter(party == "DEMOCRAT") %>% 
  mutate(county_name = str_to_title(county_name),
         vote_share = candidatevotes/totalvotes) %>% 
  drop_na()

unemp <- unemp %>% 
  cSplit(., "area_name", sep=",")
unemp <- unemp %>% 
  mutate(county_name = str_replace(unemp$area_name_1, " County", ""))
unemp <- unemp[c(1:nrow(unemp)), c(1:2,4:5)]
unemp <- unemp %>% 
  rename(state_po = area_name_2)

pov <- pov %>% 
  mutate(county_name = str_replace(pov$county_name, " County", "")) %>% 
  subset(StrRight(pov$county_name, n = 7) == " County")
pov <- pov %>% 
  rename(state_po = state)
```


## Merge data

```{r}
dem_elec_pov_merge <-
  left_join(x = dem_elec,
            y = pov,
            by.x = c("year", "state_po","county"),
            by.y = c("year","state_po","county"))

data <- 
  left_join(x = dem_elec_pov_merge,
            y = unemp,
            by.x = c("year", "state_po", "county"),
            by.y = c("year", "state_po", "county"))
data <- data %>% drop_na()
```


## Transform data

```{r}
data <- data %>% mutate(dem_maj = ifelse(vote_share > 0.50,1,0),
                        state = as.factor(state), party = as.factor(party),
                        poverty = as.integer(poverty), 
                        median_income = as.integer(median_income))
```


## Summary statistics

```{r}
data %>% summary()
```


## U.S. Presidential Election Votes by Party

```{r}
ggplot(data %>% group_by(year, party), aes(x = year, y = totalvotes, fill = party)) + 
  geom_bar(position = "stack", stat = "identity") + 
  theme_minimal() + 
  labs(title = "U.S. Presidential Election Votes by Party") + 
  xlab("Year") + ylab("Total Votes")
```


## Distribution of Democratic Vote Share for all Counties Observed Across all Elections

```{r}
hist(data$vote_share)
```


## U.S. Presidential Votes by State

```{r}
ggplot(data %>% group_by(year, state), aes(x = year, y = totalvotes, fill = state)) + 
  geom_bar(position = "stack", stat = "identity") + 
  theme_minimal() + 
  labs(title = "U.S. Presidential Election Votes by State") + 
  xlab("Year") + ylab("Total Votes")
```


## Split data (train and test)

```{r}
data_split <- initial_split(data)
data_train <- training(data_split)
data_test <- testing(data_split)
```


## Linear Regression - Predicting Democratic Vote Share Within aCounty

```{r}
mod1 <- lm(candidatevotes ~ state + poverty + median_income + unemp_rate,
           data = data_train)
summary(mod1)
```


## Linear Model Predictions

```{r}
preds_train <- predict(mod1, newdata = data_train)
summary(preds_train)

preds_test <- predict(mod1, newdata = data_test)
summary(preds_test)
```


## Linear Model RMSE

```{r}
get_rmse <- function(true, predictions){
  sqrt(mean((true - predictions)^2))}

get_rmse(data_train$candidatevotes, preds_train)

get_rmse(data_test$candidatevotes, preds_test)
```


## Linear Model MAE

```{r}
get_mae <- function(true_values, predictions){
  mean(abs(true_values - predictions))}

get_mae(data_train$candidatevotes, preds_train)

get_mae(data_test$candidatevotes, preds_test)
```


## Logistic Regression - Odds of Democratic majority in a county

```{r}
logit_dem <- glm(dem_maj ~ totalvotes + poverty + 
                   median_income + unemp_rate,
                 family = binomial,
                 data = data_train)
```
 

## Logistic coefficients

```{r}
party_odds <- data.frame("Log Odds" = logit_dem$coefficients %>% round(10), 
                         "Odds" = ((exp(logit_dem$coefficients) - 1) * 100) %>% round(10))
party_odds
```


## Confusion Matrix

```{r}
dem_scores <- predict(logit_dem, newdata = data_test, type = "response")

dem_results_logit <- data.frame(
  `truth` = as.factor(data_test$dem_maj),
  `Class1` = dem_scores,
  `predicted` = as.factor(ifelse(dem_scores > 0.6, "1", "0"))
)

dem_cm <- conf_mat(dem_results_logit,
               truth = truth,
               estimate = predicted)
print(dem_cm)
```


## Observing predicted probabilities

```{r}
hist(dem_scores)
```


## ROC Curve

```{r}
dem_roc <- ggplot(dem_results_logit, 
            aes(m = Class1, d = data_test$dem_maj)) + 
  geom_roc(labelsize = 3.5, 
           cutoffs.at = 
             c(0.6,0.4,0.2,0)) +
  geom_abline(xintercept = 0, linetype = "dashed", slope = 1) + 
  theme_minimal(base_size = 16)
calc_auc(dem_roc)
print(dem_roc)
```

## Finding Optimal 'mtry'

```{r}
data(data, package="MASS")
data_mtry <- tuneRF(data[,-17], data[,17], stepNumeric=1.5)
```


## Random Forest Model

```{r}
rf_fit_dem <- randomForest(dem_maj ~ poverty + median_income + unemp_rate + candidate + state + county_name,
                        data = data,
                        type = regression,
                        mtry = 10,
                        ntree = 500,
                       importance = TRUE )


print(rf_fit_dem)
```


## Importance Plot

```{r}
importance(rf_fit_dem,type=1, scale=FALSE)
```


## Refined Random Forest Model

```{r}
rf_fit_dem1 <- randomForest(dem_maj ~ poverty + unemp_rate + candidate + state,
                        data = data,
                        type = regression,
                        mtry = 10,
                        ntree = 500,
                       importance = TRUE )
plot (rf_fit_dem1)
```


## Refined Variable Importance Plot

```{r}
varImpPlot(rf_fit_dem1,type=1)
```


## Prediction Interaction Plot for 'unemp_rate' and 'poverty'

```{r}
library('randomForestExplainer')
plot_predict_interaction(rf_fit_dem1, 
                        data,
                         "unemp_rate",
                         "poverty")
```
